Restrictive Clearance Comparison

### 2025-01-08 data-raw/RestrictiveClearance_comparison.Rmd
Changes to Restrictive Clearance analysis document

```{r library_loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(tidyverse)
library(ctxR)
my_pk <- readRDS("data-raw/all_cvt_pk.rds")
source("data-raw/RC_comparison_bkgd.R") # httk loaded here
```

@@

@@

my_tkstats <- eval_tkstats(my_pk) %>%
  filter(Media %in% "plasma") %>%
  dplyr::select(Chemical, Species, CLtot.tkstats, CLtot.nca) %>%
  distinct() %>%
  filter(
    Species %in% "rat",
    Chemical %in% rat_pars_rat$clearance_df$Chemical,
    !is.na(CLtot.tkstats), !is.na(CLtot.nca)
  ) %>%
  group_by(Chemical, Species, CLtot.tkstats) %>%
  summarize(
    CLtot.nca_mean = mean(CLtot.nca),
    CLtot.nca_sd = sd(CLtot.nca)
  )

# Combine these two
combined_clearance_df <- my_tkstats %>% inner_join(rat_pars_rat$clearance_df)
```

@@

```{r clearance_viz}
clearance_guess <- combined_clearance_df %>%
  mutate(AD_restrictive = abs(CLtot.tkstats - restrictive_clearance),
         AD_nonrestrictive = abs(CLtot.tkstats - nonrestrictive_clearance),
         rough_determination = ifelse(
           min(AD_restrictive, AD_nonrestrictive) == AD_restrictive &
             CLtot.tkstats < nonrestrictive_clearance,
           "Restrictive", "Possibly non-Restrictive"
         )) %>% 
  dplyr::ungroup()
clearance_guess %>%
  dplyr::distinct(Chemical, rough_determination) %>%
  group_by(rough_determination) %>%
  tally()

# Preliminary viz, plot all clearances
ccd <- combined_clearance_df %>%
  relocate(CLtot.nca_mean, CLtot.nca_sd, .after = tidyselect::last_col()) %>%
  pivot_longer(
    cols = 3:6,
    names_to = "Clearance_Source",
    values_to = "Value"
  ) %>%
  mutate(
    Chemical = factor(Chemical),
    Clearance_Source = factor(
      Clearance_Source,
      levels = c(
        "CLtot.tkstats", "CLtot.nca_mean",
        "nonrestrictive_clearance", "restrictive_clearance"
      )
    )
  ) %>%
  transform(
    Chemical = fct_reorder(Chemical, Value, max)
  ) %>% 
  as_tibble()

ccd %>%
  ggplot(aes(
    x = Value,
    y = Chemical,
    fill = Clearance_Source
  )) +
  geom_point(shape = 21, alpha = 0.7) +
  scale_x_log10() +
  scale_fill_manual(
    breaks = c(
      "nonrestrictive_clearance", "CLtot.tkstats",
      "CLtot.nca_mean",
      "restrictive_clearance"
    ),
    labels = c(
      "non-restrictive clearance (httk)",
      "predicted clearance (invivoPKfit)",
      "mean non-compartmental total clearance (invivoPKfit)",
      "restrictive clearance (httk)"
    ),
    values = c("orange2", "grey5", "skyblue3", "green4")
  ) +
  facet_grid(rows = vars(Species), scales = "free", space = "free") +
  labs(
    x = "Total Clearance",
    y = "Chemical",
    title = "Comparing clearance estimates of chemicals with rat CvT data"
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
    panel.background = element_blank(),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.5),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold", angle = 0),
    # axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical",
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

@@

ggsave(
  filename = paste0(
    Sys.getenv("FIG_DIR"),"RestrictiveClearance_Analysis/",
    today,
    "rat_tkClearance_comparison_httk_PBPK.png"
  ),
  height = 8,
  width = 8,
  device = "png"
)
```

@@

```{r fup_comparison}
fup_df <- data.frame(
  Chemical = get_cheminfo("DTXSID", species = "Rat"),
  fup = as.numeric(vapply(get_cheminfo("Funbound.plasma", species = "Rat"),
                          gsub, pattern = ",.+", replacement = "",
                          FUN.VALUE = character(1)
  ))
)

fup_df %>%
  ggplot(aes(x = fup)) +
  geom_histogram(
    # aes(y = after_stat(density)),
    binwidth = 0.03,
    color = "grey5", fill = "grey5"
  ) +
  geom_histogram(
    data = fup_df %>%
      filter(Chemical %in% rat_pars_rat$param_df$Chemical),
    # aes(y = after_stat(density)),
    fill = NA,
    color = "darkgreen", binwidth = 0.03
  ) +
  expand_limits(expand = c(0,0)) +
  # geom_vline(xintercept = 0.05, linetype = "dashed") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.01))) +
  labs(
    title = "Distribution of Funbound.plasma values",
    subtitle = "(chemicals fit by invivoPKfit in dark green)"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw()

@@

test_fup <- fup_df %>%
  # filter(Chemical %in% common_chems_dawson) %>%
  pull(fup)
for (i in seq(0, 0.5, by = 1e-9)) {
  if (sum(test_fup > i) == sum(test_fup <= i)) {
    print(i)
    break
  }
}
```

@@

```{r running-invivoPKfit}
# This is now updated to use dawson chem table
minimal_pk <- pk(data = cvt %>%
                   filter(analyte_dtxsid %in% rat_pars_rat$param_df$Chemical,
                          species %in% "rat"))

new_pk <- minimal_pk +
  facet_data(vars(Chemical, Species)) +
  settings_preprocess(
    keep_data_original = TRUE,
    suppress.messages = FALSE
  ) +
  settings_optimx(method = c("L-BFGS-B")) +
  scale_conc(dose_norm = TRUE, log10_trans = TRUE) +
  stat_error_model(error_group = vars(Chemical, Species)) +
  stat_model(model = c(
    "model_1comp", "model_1comp_cl_rest", 
    "model_1comp_cl_nonrest", "model_1comp_fup",
    "model_flat")) +
  settings_optimx(method = c("bobyqa"))

@@

unique(new_pk$prefit$par_DF$param_name)

new_pk$prefit$par_DF %>%
  arrange(Chemical, Species) %>%
  filter(param_name == "Fup") %>%
  View()

new_pk <- do_fit(new_pk, n_cores = 12)
gc()

@@

@@

rel_ll %>%
  rowwise() %>% 
  mutate(across(contains("_1comp"),
                \(x) {
                  ifelse(
                    x < model_flat | x < 1,
                    NA_real_,
                    x
                  )
                })) %>% 
  # drop_na(model_1comp_fup) %>%
  filter(Chemical %in% clear_chem_levels) %>% View()
  glimpse()

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  ggplot(aes(x = estimate, y = start)) +
  geom_point(size = 2) +
  theme_bw() +
  coord_equal(xlim = c(0,1), ylim = c(0,1),
              expand = TRUE)

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  distinct(Chemical, Species, method, model,
           param_name, estimate, start) %>%
  mutate(direction = ifelse(abs(estimate - start) < 0.1,
                            "similar (less than 10% difference)",
                            ifelse(estimate > start,
                                   "increase",
                                   "decrease"))) %>%
  filter(Chemical %in% clear_chem_levels, direction %in% "similar (less than 10% difference)")
  pivot_longer(cols = c(estimate, start),
               names_to = "parameterization",
               values_to = "value") %>%
  mutate(parameterization = factor(parameterization,
                                   levels = c("start", "estimate"))) %>%
  ggplot(aes(x = parameterization, y = value)) +
  geom_point() +
  geom_line(aes(group = Chemical)) +
  facet_grid(cols = vars(direction)) +
  labs(title = "Optimization of fraction unbound in plasma") +
  theme_bw()

ggsave(filename = paste0(
  Sys.getenv("FIG_DIR"), "RestrictiveClearance_Analysis/",
  today, "fup_optimization_generalFits.png"),
  height = 5, width = 7)

@@

```{r kelim-questions}
# Original data, filtered for common_chems
# par_DF
my_kelim_prefit <- my_pk$prefit$par_DF %>%
  filter(
    param_name %in% "kelim",
    Chemical %in% common_chems_dawson
  ) %>%
  distinct(Chemical, Species, param_name, lower_bound, upper_bound, start)

my_kelim_fit <- coef(my_pk) %>%
  inner_join(get_winning_model(my_pk)) %>%
  ungroup() %>%
  distinct(model, Chemical, Species, kelim) %>%
  rename(fit_kelim = "kelim") %>%
  filter(Chemical %in% common_chems_dawson)

@@

ggsave(
  filename = paste0(Sys.getenv("FIG_DIR"), "kelim_compare_2024_0329.png"),
  height = 11,
  width = 8.5
)
```

>>>>

```{r library_loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(tidyverse)
library(ctxR)
library(ggvenn)
source("data-raw/RC_comparison_bkgd.R") # httk loaded here
# Only use plasma 
this_pk <- pk(cvt) +
  settings_optimx(method = "bobyqa") +
  facet_data(vars(Chemical, Species)) +
  settings_preprocess(
    keep_data_original = TRUE,
    suppress.messages = FALSE
  ) +
  scale_conc(dose_norm = TRUE, log10_trans = TRUE) +
  stat_error_model(error_group = vars(Chemical, Species))
this_pk <- do_fit(this_pk, n_cores = 12)
```

@@

Note: There is about 50 chemicals with rat data for both, but 3X that when
we allow substitution of human numbers. However, this does not necessarily mean
that we will get appropriate fits with the human numbers.
The aforementioned set of rat data, for example, 33 of 51 chemicals find
appropriate fits with invivoPKfit + httk.

```{r selecting_chemical_params}
chosen_list <- rat_pars_human
# The following are derived from the above selected list (see environment)
this_chem_v <- unique(chosen_list$param_df$Chemical)
this_param_list <- chosen_list$named_params
this_param_df <- chosen_list$param_df
this_httk_clearance_df <- chosen_list$clearance_df
this_species_v <- c("rat", "human")
```

We first want to see how many chemicals are in either _httk_ or _invivoPKfit_

```{r venn_diagrams}
member_list <- list(
  httk_gas_pbtk = unique(get_cheminfo(info = "DTXSID",
                                species = "human",
                                model = "gas_pbtk",
                                median.only = TRUE,
                                physchem.exclude = TRUE,
                                suppress.messages = TRUE)) %>% 
    suppressWarnings(),
  cvt = cvt %>% filter(species %in% c("human", "rat")) %>% pull(analyte_dtxsid)
)

ggvenn(member_list, fill_color = c("white", "white"))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         today, "httk_vs_CvT_VennDiagram_human.png"),
       bg = "white",
       height = 4, width = 6
)

member_list <- list(
  httk_gas_pbtk = unique(get_cheminfo(info = "DTXSID",
                                species = "rat",
                                model = "gas_pbtk",
                                median.only = TRUE,
                                physchem.exclude = TRUE,
                                suppress.messages = TRUE)) %>% 
    suppressWarnings(),
  cvt = cvt %>% filter(species %in% c("rat")) %>% pull(analyte_dtxsid)
)

ggvenn(member_list, fill_color = c("white", "white"))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         today, "httk_vs_CvT_VennDiagram_ratOnly.png"),
       bg = "white",
       height = 4, width = 6
)

new_chems <- clipr::read_clip()
length(unique(new_chems %in% setdiff(member_list$cvt, member_list$httk_gas_pbtk)))
length(unique(new_chems %in% member_list$cvt))

intersect(setdiff(member_list$cvt, member_list$httk_gas_pbtk), new_chems)
```

@@

my_tkstats <- eval_tkstats(my_pk, finite_only = FALSE) %>%
  filter(Media %in% "plasma",
         Chemical %in% this_chem_v,
         Species %in% this_species_v) %>%
  semi_join(get_winning_model(my_pk)) %>%
  dplyr::select(Chemical, Species, CLtot.tkstats, CLtot.nca) %>%
  distinct() %>%
  filter(
    !is.na(CLtot.tkstats), !is.na(CLtot.nca)
  ) %>%
  group_by(Chemical, Species, CLtot.tkstats) %>%
  summarize(
    CLtot.nca_mean = mean(CLtot.nca),
    CLtot.nca_sd = sd(CLtot.nca)
  )

# Combine these two
combined_clearance_df <- this_httk_clearance_df %>%
  inner_join(my_tkstats) %>%
  distinct()

length(unique(combined_clearance_df$Chemical))
```

@@

```{r clearance_viz}
clearance_guess <- combined_clearance_df %>%
  mutate(AD_restrictive = abs(CLtot.tkstats - restrictive_clearance),
         AD_nonrestrictive = abs(CLtot.tkstats - nonrestrictive_clearance),
         rough_determination = ifelse(
           min(AD_restrictive, AD_nonrestrictive) == AD_restrictive &
             CLtot.tkstats < nonrestrictive_clearance,
           "Restrictive", "Possibly non-Restrictive"
         )) %>% 
  dplyr::ungroup()

clearance_guess %>%
  dplyr::distinct(Chemical, rough_determination) %>%
  group_by(rough_determination) %>%
  tally()

# Preliminary viz, plot all clearances
ccd <- clearance_guess %>%
  relocate(CLtot.nca_mean, CLtot.nca_sd, .after = tidyselect::last_col()) %>%
  pivot_longer(
    cols = tidyselect::all_of(c("CLtot.tkstats", "CLtot.nca_mean",
                              "restrictive_clearance", "nonrestrictive_clearance")),
    names_to = "Clearance_Source",
    values_to = "Value"
  ) %>%
  mutate(
    Chemical = factor(Chemical),
    Clearance_Source = factor(
      Clearance_Source,
      levels = c(
        "CLtot.tkstats", "CLtot.nca_mean",
        "nonrestrictive_clearance", "restrictive_clearance"
      )
    ),
    label_cols = ifelse(rough_determination == "Restrictive",
                     "green4", "orange3")
  ) %>%
  transform(
    Chemical = fct_reorder(Chemical, Value, max)
  ) %>% 
  as_tibble()

ccd %>%
  ggplot(aes(
    x = Value,
    y = Chemical,
    fill = Clearance_Source
  )) +
  geom_point(shape = 21, alpha = 0.7) +
  scale_x_log10() +
  scale_fill_manual(
    breaks = c(
      "nonrestrictive_clearance", "CLtot.tkstats",
      "CLtot.nca_mean",
      "restrictive_clearance"
    ),
    labels = c(
      "non-restrictive clearance (httk)",
      "predicted clearance (invivoPKfit)",
      "mean non-compartmental total clearance (invivoPKfit)",
      "restrictive clearance (httk)"
    ),
    values = c("orange2", "grey5", "skyblue3", "green4")
  ) +
  facet_grid(rows = vars(Species), scales = "free", space = "free") +
  labs(
    x = "Total Clearance",
    y = "Chemical",
    title = "Comparing clearance estimates of chemicals with rat CvT data"
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
    panel.background = element_blank(),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.5),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold", angle = 0),
    axis.title.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical",
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

@@

ggsave(
  filename = paste0(
    Sys.getenv("FIG_DIR"),"RestrictiveClearance_Analysis/",
    today,
    "all_tkClearance_comparison_httk_PBPK.png"
  ),
  height = 8,
  width = 8,
  device = "png"
)
```

@@

```{r fup_comparison}
fup_df <- this_param_df %>%
  distinct(Chemical, Species, forced_human_values,
           Funbound.plasma, Funbound.plasma.adjustment, 
           Funbound.plasma.dist) %>%
  mutate(
    Funbound.plasma.median = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "^\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.lower95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.upper95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d$"
    ))
    )

fup_df %>%
  ggplot(aes(x = Funbound.plasma.adjustment)) +
  geom_histogram(
    data = fup_df %>%
      filter(Chemical %in% this_httk_clearance_df$Chemical),
    # aes(y = after_stat(density)),
    fill = "white",
    color = "darkgreen", binwidth = 0.03
  ) +
  expand_limits(expand = c(0,0)) +
  geom_vline(xintercept = 0.05, linetype = "dashed") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.01))) +
  labs(
    title = "Distribution of Funbound.plasma values",
    subtitle = "(chemicals fit by invivoPKfit in dark green)"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw()

@@

test_fup <- fup_df %>%
  pull(Funbound.plasma)
for (i in seq(0, 0.5, by = 1e-9)) {
  if (sum(test_fup > i) == sum(test_fup <= i)) {
    print(i)
    break
  }
}

fup_df %>%
  ggplot(aes(x = Funbound.plasma, y = Funbound.plasma.adjustment)) +
  geom_point()

@@

```{r running-invivoPKfit}
# This is now updated to use dawson chem table
minimal_pk <- pk(data = cvt %>%
                   filter(analyte_dtxsid %in% this_chem_v,
                          species %in% "rat",
                          administration_route_normalized == "iv"))

new_pk <- minimal_pk +
  facet_data(vars(Chemical, Species)) +
  settings_preprocess(
    keep_data_original = TRUE,
    suppress.messages = FALSE
  ) +
  scale_conc(dose_norm = TRUE, log10_trans = TRUE) +
  stat_error_model(error_group = vars(Chemical, Species)) +
  stat_model(model = c(
    "model_1comp", "model_1comp_cl_rest", 
    "model_1comp_cl_nonrest", "model_1comp_fup",
    "model_flat")) +
  settings_optimx(method = c("bobyqa"))

@@

unique(new_pk$prefit$par_DF$param_name) # Check all parameters are there
length(unique(new_pk$data$Chemical)) 
length(unique(this_chem_v))

new_pk <- do_fit(new_pk, n_cores = 14)
gc()

@@

new_tkstats <- eval_tkstats(new_pk, finite_only = TRUE)

@@

rel_ll %>%
  rowwise() %>% 
  mutate(across(contains("_1comp"),
                \(x) {
                  ifelse(
                    x < model_flat | x < 1,
                    NA_real_,
                    x
                  )
                })) %>% 
  count(is.na(model_1comp_fup))
  # drop_na(model_1comp_fup) %>%
  # filter(Chemical %in% clear_chem_levels) %>% View()
  glimpse()

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  ggplot(aes(x = estimate, y = start)) +
  geom_point(size = 1.25) +
  geom_point(data = new_pk$fit %>%
               filter(param_name %in% "Fup",
                      model %in% "model_1comp_fup",
                      Chemical %in% combined_clearance_df$Chemical),
             size = 1.5, color = "green4") +
  theme_bw() +
  coord_equal(xlim = c(0,1), ylim = c(0,1),
              expand = TRUE) +
  coord_flip()

ggsave(filename = paste0(
  Sys.getenv("FIG_DIR"), "RestrictiveClearance_Analysis/",
  today, "fup_optimization_allowingHuman.png"),
  height = 5, width = 5)

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  distinct(Chemical, Species, method, model,
           param_name, estimate, start) %>%
  mutate(direction = ifelse(abs(estimate - start) < 0.1,
                            "similar (less than 10% difference)",
                            ifelse(estimate > start,
                                   "increase",
                                   "decrease"))) %>%
  # filter(direction %in% "similar (less than 10% difference)") %>%
  pivot_longer(cols = c(estimate, start),
               names_to = "parameterization",
               values_to = "value") %>%
  mutate(parameterization = factor(parameterization,
                                   levels = c("start", "estimate"))) %>%
  ggplot(aes(x = parameterization, y = value)) +
  geom_point() +
  geom_line(aes(group = Chemical)) +
  facet_grid(cols = vars(direction)) +
  labs(title = "Optimization of fraction unbound in plasma") +
  theme_bw()

ggsave(filename = paste0(
  Sys.getenv("FIG_DIR"), "RestrictiveClearance_Analysis/",
  today, "fup_optimization_generalFits_allowingHuman.png"),
  height = 5, width = 7)

@@

```{r kelim-questions}
# Original data, filtered for common_chems
# par_DF
my_kelim_prefit <- my_pk$prefit$par_DF %>%
  filter(
    param_name %in% "kelim",
  ) %>%
  distinct(Chemical, Species, param_name, lower_bound, upper_bound, start)

my_kelim_fit <- get_fit(my_pk) %>%
  semi_join(get_winning_model(my_pk)) %>%
  ungroup() %>%
  filter(param_name == "kelim") %>%
  distinct(model, Chemical, Species, param_name, value)

@@

ggsave(
  filename = paste0(Sys.getenv("FIG_DIR"), "kelim_compare.png"),
  height = 11,
  width = 8.5
)
```
